# Reproducible Research: Peer Assessment 1

The activity data was collected by an anonymous individual wearing a personal activity monitoring device 
for two months (from 2012-10-01 to 2012-11-30) and the number of steps was recorded in 5-minutes intervals.

First, we make sure that all code chunks are visible:
```{r setoptions, echo=TRUE}
opts_chunk$set(echo=TRUE)
```

## Loading and preprocessing the data

The data file `activity.csv` is extracted from the compressed file `activity.zip` and then loaded into the data frame `data`.
The interval data are transformed from the orignial format (hhmm, no digits for leading zeros) 
into minutes passed since the beginning of the day (00:00).
In this way the intervals are spread evenly over the day 
(and gaps in the original data like 50, 55, 100, 105 are transformed into 50, 55, 60, 65).
The structure of the data frame and a summary statistic are displayed.
```{r loaddata}
data <- read.csv(unz("activity.zip", "activity.csv"), colClasses=c("integer", "Date", "integer"))

data$interval <- 60*floor((data$interval+1)/100) + (data$interval %% 100)

str(data)
summary(data)
```


## What is mean total number of steps taken per day?

The total steps per day are summed up using the `tapply` function, and the mean and median are determined.
```{r meansteps}
total_steps <- tapply(data$steps, data$date, sum, na.rm=T)

step_mean <- mean(total_steps)
step_mean
step_median <- median(total_steps)
step_median
```

The total steps per day are displayed as a histogram.
The mean value of the total number of steps taken per day (`r round(step_mean, 2)`) is highlighted by a vertical red line, 
the median (`r round(step_median, 2)`) by a vertical blue line.
The mean is shifted to the left relative to the median.
```{r histogram}
hist(total_steps, breaks=11, 
     xlab="number of steps per day", 
     main="Histogram of total steps per day")
abline(v=step_mean, col="red", lwd=3)
abline(v=step_median, col="blue", lwd=3)
legend(x="topright", legend=c("mean","median"), col=c("red","blue"), bty="n", lwd=3)
```


## What is the average daily activity pattern?

To generate an average daily activity pattern the mean of each 5-minutes interval over all days is determined using the `tapply` function. 
The activity pattern is plotted as a time series.
```{r timeseries, fig.height=5}
avg_steps <- tapply(data$steps, data$interval, mean, na.rm=T)

hours <- as.numeric(names(avg_steps))/60
plot(hours, avg_steps, type="l", axes=F,
     xlab="time of day (h)", ylab="average number of steps in 5-min interval",
     main="Daily activity pattern")
axis(2)
axis(1, at=0:6*4, labels=paste(0:6*4,":00", sep=""))
```

The maximum number of steps occurs in the 5-minutes interval starting at
```{r maxactivity}
max_act_num <- which(avg_steps==max(avg_steps))
max_act_int <- data$interval[max_act_num]
sprintf("%02d:%02d", floor(max_act_int/60), max_act_int %% 60)
```

The maximum number of steps occurs in the `r max_act_num`th 5-minutes interval of the day 
starting at `r paste(floor(max_act_int/60), ":", max_act_int %% 60, sep="")`.

## Imputing missing values
There are many missing values in the data set (see the data summary statistic at the top of this document), to be exact `r sum(is.na(data))` 
missing values:
```{r missingvalues}
sum(is.na(data))
```

The daily activity pattern can be used to impute these missing values. For every missing value in the orignial data set 
the average number of steps in that 5-minutes interval is used and a new data frame `impute` is created. 
This procedure should be valid if the person has daily routine, i.e. an activity pattern that is similar over multiple days.
Instead of missing values this data set now contains a typical value of that 5-minutes interval.
```{r imputation}
impute <- transform(data, steps=ifelse(is.na(steps), avg_steps, steps))

summary(impute)
```

Now, using the data set with imputed values, the total steps per day are again summed up using the `tapply` function, 
and the mean and median are determined.
```{r meanimpsteps}
total_impsteps <- tapply(impute$steps, impute$date, sum, na.rm=T)

impstep_mean <- mean(total_impsteps)
impstep_mean
impstep_median <- median(total_impsteps)
impstep_median
```

The total steps per day are displayed as a histogram.
The mean value of the total number of steps taken per day (`r round(impstep_mean, 2)`) is highlighted by a vertical red line, 
the median (`r round(impstep_median, 2)`) by a vertical blue line.
The mean and the median overlap, and the peak of days with no recorded steps is gone.
Both values have increased compared to the original data set. The increase of the mean, however, is much stronger.
```{r imphistogram}
hist(total_impsteps, breaks=11, 
     xlab="number of steps per day", 
     sub="(missing values imputed)",
     main="Histogram of total steps per day")
abline(v=impstep_mean, col="red", lwd=3)
abline(v=impstep_median, col="blue", lwd=3, lty=2)
legend(x="topright", legend=c("mean","median"), col=c("red","blue"), bty="n", lwd=3)
```

Due to imputation the total sum of steps in these two month increases from `r sum(data$steps, na.rm=TRUE)` to `r sum(impute$steps)`.
```{r}
sum(data$steps, na.rm=TRUE)

sum(impute$steps)
```


## Are there differences in activity patterns between weekdays and weekends?
In order to identify differences between weekdays and weekends a daily acitity pattern is generated for both types of days.
First the data is classified as recorded either on a weekday or on a weekend, and this information is stored in `week`.
Then the data is aggregated by 5-minutes interval and weekday/-end.
Using the `ggplot` function a panel plot contrasting the weekday and weekend activity is produced.

```{r weekends}
week <- factor(weekdays(impute$date) %in% c("Saturday","Sunday"), 
               labels=c("weekday","weekend"), ordered=FALSE)

impsteps <- aggregate(impute$steps, by=list(interval=impute$interval, weekday=week), mean)

library(ggplot2)
g <- ggplot(impsteps, aes(interval/60, x))
g + geom_line() + facet_grid(weekday ~ .) +
    scale_x_continuous(breaks=0:6*4, labels=paste(0:6*4,":00", sep="")) +
    theme_bw() +
    labs(y="average number of steps in 5-min interval") +
    labs(x="time of day (h)") +
    labs(title="Daily activity pattern")
```

Whereas peak and morning activity is highest during weekdays, the overall activity is higher on weekends.
In light of the differences between weekdays and weekends it might be a better imputation strategy to use 
weekend interval averages for imputing weekend activity and weekday interval averages for imputing weekday activity.